---
title: "dplyr vs apply functions"
author: "William Surles"
date: "February 15, 2016"
output: 
  html_document:
    self_contained: yes
    theme: flatly
---

This document examines the apply functions as taught in the datacamp course 
'Intermediate R' and how to complete the exercises with dplyr and other libraries.

I think these lessons are great for learning the basics of apply functions.
However, Hadley Wickham has created many R libraries that allow an analyst to 
do this same work with much cleaner code, like dplyr and purrr.
There are also libraries that make working with strings more consistent, like stringr.

This is meant to be a companion to the apply function exercises in the Intermediate R
class. 

Why are these libraries better?

  - First, becuse they have more consistent function names so it will be easier for 
you to remember them. 
  - Second, they can chain and manipulate dataframes or lists in a much clearner code format.
    - Its one chunck of code that an analyst can read in sequence, instead of many separate chuncks
of code where you have to assign the outcome to new variable names. 
    - Having to assign your outocme to a variable at each step takes extra thought for the developer to make names or gets messy if its var1, var2, var3. 
    - A future analyst also has to look through all the code, not intuitivelly knowing what 
varibales are just temporary and which are actually important. 
    - With dply and purr you can do all the steps in one chunck and then assign the outcome to one variable which is important.

First, lets load some libraries that we will use here.

```{r, message=F}
library(dplyr)
library(purrr)
library(stringr)
library(ggplot2)
```

## "Use lapply with a built-in R function" Result
Lets look at the first exercise with lapply. 
This is the correct code for the exercise.

```{r, cache=T}
# The vector pioneers has already been created for you
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

# Split names from birth year: split_math
split_math <- strsplit(pioneers, ":")

# Convert to lowercase strings: split_low
split_low <- lapply(split_math, tolower)

# Take a look at the structure of split_low
str(split_low)

```

Thats fine and all, but it would be really nice if we could just do this all in 
one code chunk

## Using purrr and stringr
This is how I would redo it. 

```{r, cache=T}

## This is cool and stringr is a great package for working with strings
## It does not however acieve what I want
## Wait actually it does because it splits the string
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")

split_low <- pioneers %>% 
  str_split(":") %>% 
  map(str_to_lower)

str(split_low)

```

First, I am using [stringr](https://cran.r-project.org/web/packages/stringr/stringr.pdf) 
functions. They are more consistent and I recommend 
checking out and using this package for string manipulations. 

Second, using [purrr](https://github.com/hadley/purrr) makes this way better. 
purrr allows me to use chains, so the output of the first function can be passed 
to the next easily. Its assumed as the first variable. 
This way the analyst can see you are starting with the pioneers vector and then 
performing functions on it and they can easily read left to right or top to bottom.
They do not need to go find the start of the function string somewhere in the middle
of the code glob. Its simple here but I have often seen many functions nested together.
Its so hard to easily read whats happening that way. 

The map function works like the lapply. 
It takes the object passed through the chain as the first argument (the list we 
just split) and applies the function that we specify to all elements in the list. 

## Use lapply with your own function Result

```{r}
# Code from previous exercise:
pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split <- strsplit(pioneers, split = ":")
split_low <- lapply(split, tolower)

# Write function select_first()
select_first <- function(x) {
  return(x[1])
}

# Apply select_first() over split_low: names
names <- lapply(split_low, select_first)

# Write function select_second()
select_second <- function(x) {
  return(x[2])
}

# Apply select_second() over split_low: years
years <- lapply(split_low, select_second)
```

## Redo Using purrr and stringr
This is how I would redo it. 

```{r, cache=T}

pioneers <- c("GAUSS:1777", "BAYES:1702", "PASCAL:1623", "PEARSON:1857")
split_low <- pioneers %>% str_split(":") %>% map(str_to_lower)

names <- split_low %>% map(function(x) x[1])
years <- split_low %>% map(function(x) x[2])
```

With purrr you can pass functions directly to the map. In this case, where we
just want to pull out a specified element this makes a lot of sense and its nicer
to have the function right there so you can see whats happening. 
But if it were a more complex function than you could definite it separately and 
call it the same way we call str_to_lower

However, I don't really like any of this because its taking someting simple
and making it harder by putting it into a list. 
A better way to do this altogether is to use str_split_fixed (in stringr) 
to put the result into columns instead of a list. 
Then you can put the values into a data frame and run any functions you want 
on the individual columns. 

In the exercise we are actually running tolower on the numbers because it thinks
they are characters. It does not do anything here, but in other cases 
you many not want to run a function on all the values in the list. Having
the data split into a dataframe first thing will make manipulation much more
straight forward and easier to control.


```{r, cache=T}
df <- pioneers %>% 
  str_split_fixed(":", n=2) %>%
  data.frame() %>%
  mutate(
    names = str_to_lower(as.character(X1)), 
    years = as.numeric(X2)
    ) %>%
  select(names,years)
  
str(df)
```

However, if you do have a reason to map the functions to the list then
you can pull the elements in each list into columns of a data frame with
`do.call` and `rbind` once you are finished.

```{r}
df <- data.frame(do.call('rbind', split_low))
df
```

Lets keep goin with the exercises. I'm going to go a little faster here and just
go through the solutions with purrr.

You can pass values through with the ~ symbol to an equation and reference the 
list passed through with a `.`. This is the same as the anonymous function.

```{r, cache=T}
names <- split_low %>% map(function(x) x[1])
names <- split_low %>% map(~ .[1])
```

You can also pass other function values through just like with lapply
```{r, cache=T}
select_el <- function(x,pos) {return(x[pos])}
names <- split_low %>% map(select_el, pos=1)
```

What about other things in lists that you can't split into a dataframe.
I can easily still use purrr for this. Here we can apply the min and max functions
to the temp variable with purrr. The main issue with sapply is that it may convert
the results to a vetor or it may not. This can be problematic for code. With purrr
you know that you will have a list as the output and can code accordingly.

```{r, cache=T}
## This recreates the temp car used in the sapply exercises
temp <- list(c(3,7,9,6,-1),
             c(6,9,12,13,5),
             c(4,8,3,-1,-3),
             c(1,4,7,2,-2),
             c(5,7,9,4,2),
             c(-3,5,8,9,4),
             c(3,6,9,4,1))

temp %>% map(min)
temp %>% map(max)

```



## More Resources
  - [purrr 0.1.0 article on rbloggers](http://www.r-bloggers.com/purrr-0-1-0/)
  - [stringr vignette](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html)
  - I'll cover more with purrr in this document.[link to another markdown here]

