---
title: "purr examples"
author: "William Surles"
date: "February 15, 2016"
output: 
  html_document:
    self_contained: yes
    theme: flatly
---

Libraries used

```{r, message=F}
library(purrr)
library(ggplot2)
```


## A simple purrr example
Here is a simple purrr example that splits a dataframe into a list and performs
operations on the df in each list separately

```{r, cache=T}
mtcars %>%
  split(.$cyl) %>%
  map(~ lm(mpg ~ wt, data = .)) %>%
  map(summary) %>%
  map_dbl("r.squared")

```

Wow, thats pretty cool yeah.

You can even get more complex here and create charts!

## A cooler purrr example
```{r, cache=T}
library(gridExtra)

p <- mtcars %>%
  split(.$am) %>%
  map( ~ ggplot(data=.x, aes(x=wt, y=mpg, colour=as.factor(cyl))) +
         geom_point()
       )
p
```

You can split a data frame based on a variable into separate list elements, 
then pass that to a plot function. Each df in the list elements will be ploted.
Thats very cool. 

I tried for a little bit to arange these in columns with marrangeGrobs but could 
not quite figure out how to adjust the widths. This is fine for now as an example.
